Variance reduction for MCMC methods via martingale representations

### 2020-02-17 aap/vr-mcmc.tex
update

\newcommand{\eric}[1]{\todo[color=purple!10, author=Eric]{#1}}
\newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
\newcommand{\denis}[1]{\todo[color=green!10, author=Denis]{#1}}
\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
\newcommand{\sergey}[1]{\todo[color=red!10, author=Sergey]{#1}}
\newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{T}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

%added command for indicator - SS
\newcommand{\indiacc}[1]{\mathbbm{1}_{\{#1\}}}

@@

\begin{aug}
\author{\snm{D. Belomestny}\thanksref{a,b}\ead[label=e1]{denis.belomestny@uni-due.de}}
\and
\author{\snm{E. Moulines}\thanksref{c, b}\ead[label=e2]{eric.moulines@polytechnique.edu}}
\and
\author{\snm{S. Samsonov}\thanksref{b}\ead[label=e3]{svsamsonov@hse.ru}}
\and
\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e4]{nshagadatov@edu.hse.ru}}

@@

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par
Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the so-called control variates variance reduction method
 is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
such that the variance of the r.v. $f(X)-\zeta$ is small.  The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and so-called Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the so-called Poisson equation for the corresponding  Markov chain. It was observed in Henderson~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).  The best choice for the function \(G\) corresponds to a solution of the so-called Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  Henderson~\cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which doesn't require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA). Section~\ref{sec:coeff} studies the complexity of the variance reduced ULA. Finally, numerical examples are presented in Section~\ref{sec:num}.

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
${\mu=\frac{\nabla U}{2}}$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
\end{equation}
provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

Then, as shown in Metropolis et al.~\cite{metropolis1953equation}, this transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\). If \(q_p\) have a wide enough

@@

dX_t=b(X_t)\,dt+\sigma(X_t)\,dW_t,\quad t\geq 0,	

@@

\par

@@

The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of simplicity we restrict our attention to the constant time step.
We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2\right]
&=&\left(\mathsf{E}[\rho(f)]-\pi(f)\right)^2
+\mathsf{Var}[\rho(f)].
Our analysis will be done under the following two assumptions.

>>>>

\newcommand{\eric}[1]{\todo[color=purple!10, author=Eric]{#1}}
\newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
\newcommand{\denis}[1]{\todo[color=green!10, author=Eric]{#1}}
\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Eric]{#1}}

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{T}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

@@

\begin{aug}
\author{\snm{D. Belomestny}\thanksref{a,b}\ead[label=e1]{denis.belomestny@uni-due.de}}
\and
\author{\snm{E. Moulines}\thanksref{c, b}\ead[label=e2]{eric.moulines@polytechnique.edu}}
\and
\author{\snm{N. Shagadatov}\thanksref{b}\ead[label=e3]{nshagadatov@edu.hse.ru}}
\and
\author{\snm{M. Urusov}\thanksref{a}\ead[label=e4]{murusov@uni-due.de}}

@@

Monte Carlo integration typically has an error variance of the form
$\sigma^{2}/n,$ where $n$ is a sample size and \(\sigma^2\) is the variance of  integrand. We can make the variance smaller
by using a larger value of $n$.  Alternatively,
 we can  reduce $\sigma^2$ instead of increasing the sample size \(n\). To this end, one can try to construct
a new Monte Carlo experiment with the same expectation as the original one
but with a lower variance $\sigma^2$. Methods to achieve this are known as variance
reduction techniques. Variance reduction plays an important role in
Monte Carlo and Markov Chain Monte Carlo methods. Introduction to many of the variance reduction techniques can be found in \cite{christian1999monte}, \cite{rubinstein2016simulation}, \cite{GobetBook} and \cite{glasserman2013monte}. Recently one witnessed a revival of interest in efficient variance reduction methods  for MCMC algorithms, see for example \cite{dellaportas2012control}, \cite{mira2013zero}, \cite{brosse2018diffusion} and references therein.
\par
Suppose that we wish to compute $\pi(f):=\mathsf{E}\left[f(X)\right]$, where $X$
is a random vector-valued in $\mathcal{X}\subseteq\mathbb{R}^{d}$ with a density $\pi$ and $f:$
$\mathcal{X}\to\mathbb{R}$ with $f\in L^2(\pi)$.
The idea of the control variates variance reduction method
 is to find a cheaply computable random variable $\zeta$ with $\mathsf{E}[\zeta]=0$ and \(\mathsf{E}[\zeta^2]<\infty,\)
such that the variance of the r.v. $f(X)-\zeta$ is small.  The complexity of the problem of constructing classes $Z$ of control variates \(\zeta\) satisfying   $\mathsf{E}[\zeta]=0$ essentially depends on the degree of our knowledge on \(\pi.\)
For example, if \(\pi\) is analytically known and satisfies some regularity conditions, one can apply the well-known technique of  polynomial interpolation to construct control variates enjoying  some optimality properties, see, for example, Section~3.2 in \cite{dimov2008monte}. Alternatively, if an orthonormal system in \(L^2(\pi)\) is analytically available, one can build control variates \(\zeta\) as a linear combination of the corresponding basis functions, see \cite{GobetCV}. Furthermore, if \(\pi\) is known only up to a normalizing constant (which is often the case in Bayesian statistics), one can apply the recent approach of control variates  depending only on the gradient \(\nabla \log \pi\)  using Schr\(\ddot{\text{o}}\)dinger-type Hamiltonian operator in  \cite{mira2013zero} and Stein operator in \cite{brosse2018diffusion}.
In some situations \(\pi\) is not known analytically, but \(X\) can be represented as a function of  simple random variables with known distribution.
Such  situation arises, for example, in the case of functionals of  discretized diffusion processes. In this case a Wiener chaos-type decomposition can be used to construct control variates with nice theoretical properties, see \cite{belomestny2018stratified}.
Note that in order to compare different  variance reduction approaches, one has to analyze their complexity, that is, the number of numerical operations required to achieve a prescribed magnitude of the resulting variance.

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain. It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).  The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA). Section~\ref{sec:coeff} studies the complexity of the variance reduced ULA. Finally, numerical examples are presented in Section~\ref{sec:num}.

\section{Setup}\label{sec:setup}
Let  \(\mathcal{X}\) be a domain in \( \mathbb{R}^d.\)  Our aim is to numerically compute  expectations of the form
\[
\pi(f)=\int_{\mathcal{X}} f(x)\pi(\rmd x),
\]
where \(f:\) \(\mathcal{X}\longrightarrow \mathbb{R}\) and \(\pi\) is a probability measure supported on \(\mathcal{X}.\)
If  the dimension of the space \(\mathcal{X}\) is large and \(\pi(f)\) can not be computed analytically, one can apply Monte Carlo methods. However, in many practical situations  direct sampling from \(\pi\) is impossible and this precludes the use of plain Monte Carlo methods in this case. One popular alternative to Monte Carlo  is Markov Chain Monte Carlo (MCMC), where one is looking for a discrete time  (possibly non-homogeneous) Markov chain   \((X_p)_{p\geq 0}\) such that \(\pi\) is its unique invariant measure. In this paper we study a class of MCMC algorithms with \((X_p)_{p\geq 0}\) satisfying the  the following recurrence relation:
\begin{equation}
\label{eq:chain_gen}
X_{p}=\Phi_{p}(X_{p-1},\xi_{p}),\quad p=1,2,\ldots ,\quad X_{0}=x_0,
\end{equation}
for some i.i.d.  random vectors \(\xi_p\in \mathbb{R}^m\) with distribution \(P_{\xi}\)
and some Borel-measurable
functions $\Phi_{p}\colon\mathcal{X}\times\mathbb{R}^{m}\to\mathcal{X}.$
In fact, this is quite general class of Markov chains (see Theorem~1.3.6 in ~\cite{moulines2018})
and many well-known MCMC algorithms can be represented in form \eqref{eq:chain_gen}.
Let us consider two popular examples.
\begin{example}[Unadjusted Langevin Algorithm]
\label{exam:langevin-algorithm}
Fix a sequence of positive time steps \((\gamma_p)_{p\geq 1}.\) Given a Borel function $\mu\colon\mathbb{R}^{d}\to\mathbb{R}^{d}$,
consider a non-homogeneous
discrete-time Markov chain $(X_{p})_{p\geq0}$ defined by
\begin{equation}\label{eq:chain}
X_{p+1}=X_{p}-\gamma_{p+1}\mu(X_{p})+\sqrt{\gamma_{p+1}}Z_{p+1},\end{equation}
where $\left(Z_{p}\right)_{p\geq1}$ is an i.i.d.\ sequence of $d$-dimensional
standard Gaussian random vectors. If
$\mu=(1/2) \nabla U$
for some continuously differentiable function $U,$ then Markov chain~\eqref{eq:chain} can be used to approximately sample from the density
\begin{equation}\label{eq:stationary_distr}
\pi(x)= \mathrm{const}\,\rme^{-U(x)}, \quad \mathrm{const}= \left. 1 \middle /{\int_{\mathbb{R}^{d}} \rme^{-U(x)}\, \rmd x,} \right.
\end{equation}
provided that \(\int_{\mathbb{R}^{d}} \rme^{-U(x)}\,\rmd x\) is finite. This method is usually referred to as
Unadjusted Langevin Algorithm (ULA).
In fact the Markov chain~\eqref{eq:chain}
arises as the Euler-Maruyama discretization
of the Langevin diffusion
\[
\rmd Y_t=-\mu(Y_t)\,\rmd t+ \rmd W_t
\]
with nonnegative time steps $(\gamma_p)_{p\ge1}$,
and,  under mild technical conditions, the latter Langevin diffusion admits $\pi$
of~\eqref{eq:stationary_distr}
as its unique invariant distribution; see \cite{dalalyan2017theoretical} and \cite{durmus:moulines:2017}.
\end{example}
\begin{example}[Metropolis-Adjusted Langevin Algorithm]
The Metropolis-Hastings algorithm
associated with a target density \(\pi\) requires the choice of a sequence of conditional densities  \((q_p)_{p\geq 1}\) also called proposal or candidate kernels. The transition from the value  of the Markov chain \(X_p\)  at time \(p\)
and its value at time \(p + 1\) proceeds via the following transition step:

@@

This transition is reversible with respect to \(\pi\) and therefore preserves the stationary density \(\pi\); see \cite[Chapter~2]{moulines2018}. If \(q_p\) have a wide enough

@@

dX_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	

@@

@@

\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. In this section   we provide  an analysis of the variance-reduced ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
We shall use the notations $\mathbb N=\{1,2,\ldots\}$ and $\mathbb N_0=\mathbb N\cup\{0\}$.
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{p=N+1}^{N+n}\omega_{p,n}^{N}\sum_{l=N+1}^{p}a_{p,l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\label{eq:29032018a5}\\
&=\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=N+1}^{N+n}\bar{a}_{l,\mathbf{k}}(X_{l-1})\mathbf{H}_\mathbf{k}(Z_{l})
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i\) and
\begin{align*}
\bar{a}_{l,\mathbf{k}}(x)
& =\sum_{p=l}^{N+n}\omega_{p,n}^{N}a_{p,l,\mathbf{k}}(x) =\mathsf{E}\left[\left.\left(\sum_{p=l}^{N+n}\omega_{p,n}^{N}f(X_{p})\right)\mathbf{H}_\mathbf{k}\left(Z_{l}\right)\right|X_{l-1}=x\right].
\end{align*}
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we are interested in its conditional
Mean Squared Error (MSE),
which can be decomposed as the sum of the squared conditional bias
and the conditional variance:
\begin{eqnarray}\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)|\mathcal G_N\right]
&=&\mathsf{E}\left[(\rho(f)-\pi(f))^2|\mathcal G_N\right]
\\
\nonumber
&=&\left(\mathsf{E}[\rho(f)|\mathcal G_N]-\pi(f)\right)^2
+\mathsf{Var}\left[\rho(f)|\mathcal G_N\right].
\end{eqnarray}
The quantities in~\eqref{eq:29032018a4}
are conditioned on $\mathcal G_N$,
as it reflects the way the estimators are used for MCMC:
the path of the Markov chain is simulated only once,
and we start to use the realized values
of the Markov chain to construct
our estimate only after the burn-in period of length~$N$.
We also notice that, due to the Markovian structure,
the conditioning on $\mathcal G_N$
in~\eqref{eq:29032018a4}
is equivalent to conditioning on $X_N$ only
(this is particularly clear in the case $\rho(f)=\pi_n^N(f)$
but requires some calculation
in the remaining case $\rho(f)=\pi_{K,n}^N(f)$).

\subsection{Squared conditional bias}
Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[\left.M_{K,n}^N(f)\right|\mathcal G_N\right]=0,
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$ For a bounded Borel function~$f$,
we can estimate the conditional bias
similarly to the beginning of ~\cite[Section~4]{durmus:moulines:2017}:
\begin{align}
\left(\mathsf{E}[\pi_{K,n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
&=\left(\mathsf{E}[\pi_{n}^N(f)|\mathcal G_N]-\pi(f)\right)^2
\label{eq:06042018a2}\\
&\le
\mathrm{osc}(f)^2
\sum_{p=N+1}^{N+n}
\omega_{p,n}^N\,
\|Q_\gamma^{N+1,p}(X_N,\cdot)-\pi(\cdot)\|_{TV}^2,
\notag
\end{align}
where
$\mathrm{osc}(f):=\sup_{x\in\mathbb R^d}f(x)-\inf_{x\in\mathbb R^d}f(x)$,
$\|\mu-\nu\|_{\TV}$ denotes the total variation distance
between probability measures $\mu$ and $\nu$, that is,
$$
\|\mu-\nu\|_{TV}=\sup_{A\in\mathcal B(\mathbb R^d)}
|\mu(A)-\nu(A)|,
$$
$\pi(\cdot)$ denotes the probability measure on $\mathbb R^d$
with density $\pi$ of~\eqref{eq:stationary_distr};
for $\gamma>0$, the Markov kernel $R_\gamma$
from $(\mathbb R^d,\mathcal B(\mathbb R^d))$
into $(\mathbb R^d,\mathcal B(\mathbb R^d))$
is defined by
$$
R_\gamma(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\}dy,
$$
while, for $k,l\in\mathbb N$, $k\le l$,
the kernel $Q_\gamma^{k,l}$ is
$
Q_\gamma^{k,l}=R_{\gamma_l}\cdots R_{\gamma_k},
$
which, finally, provides the (random) measure
$Q_\gamma^{N+1,p}(X_N,\cdot)$
used in the right-hand side of~\eqref{eq:06042018a2}.
Different specific upper bounds for the squared bias
can be deduced from~\eqref{eq:06042018a2}
using results of Section~3 in~\cite{durmus:moulines:2017}
on bounds in the total variation distance.
